Random Matrix Theory Tools for the Analysis of Functional Magnetic Resonance Imaging Examinations

### 2021-09-13 paper/latex/rmt.tex
wip

For this paper’s analyses, brain extraction is necessary to limit analyses to the appropriate
voxels, motion correction is necessary to allow proper brain extraction, and both motion-correction
and slice-time correction simplify the interpretation of voxel-voxel correlations across voxel
locations. Whereas physiological noise regression and bias field correction would perhaps be
appropriate here, there is insufficient data to perform these steps for all datasets (Table II).
Thus, motion-correction was performed with FSL's MCFLIRT [56], slicetime correction with the default
method in FSL, and brain extraction was performed with BET [57]. Due to differences in the size of
the REFLECT data, the higher-resolution resting-state data was interpolated to have the same spatial
dimensions as the task data.

@@

Two classes of features were examined in this study: those that require unfolding, and those that do
not. In order to compare the observed eigenvalue distribution to some theoretical predictions, the
eigenvalues must be unfolded [5], [6]. The details and motivation behind the unfolding procedure are
well documented in [5]. Practically, the unfolding process can be viewed as a smoothing and
rescaling procedure where the originally observed spectrum \(\lambda_1 \le \dots \le \lambda_n\) is
smoothly mapped to the unfolded spectrum \(e_1, \dots, e_n\), such that \(\sum d_i/n \approx 1\),
if \(d_i = e_{i+1} - e_i \). For most empirical data, the true smoothing function is not known, and
so must be approximated [5], [6]. Typically, a polynomial is chosen [58].

A number of statistics (or "spectral observables" [6]) can then be computed from this unfolded
spectrum. In the literature, the two most common points of comparison for these unfolded features
are typically the features of the Poisson or Gaussian Diagonal Ensemble (GDE) and the GOE (see e.g.
[12], [14], [15], [24]–[26], [32], [33]). The GOE (or GDE) consists of real, symmetric matrices
where entries (or only diagonal entries) are normal random variables with zero mean and unit
variance. When the empirically observed features are close to those of the GOE, this is usually
interpreted as indicating a high degree of randomness and statistical independence in the data, and
when the observed features are closer to those expected in the GDE or Poisson case, this is
interpreted as indicating a high degree of coherence or dependence [12], [14], [15], [24]–[26],
[32]–[34].

The Nearest-Neighbours Spacings Distribution (NNSD) is the most basic RMT feature, and is the
distribution of the spacings (differences) of adjacent unfolded eigenvalues. Normally, the NNSD is
analysed visually via a histogram of the distribution of spacings, however it is also possible to
analyse it more formally by fitting the Brody distribution [59]:

@@

Whereas the NNSD and Brody parameter characterize the local properties of eigenvalues close on the
spectrum, the long-range properties relating eigenvalues distant on the spectrum can be
characterized by both the spectral rigidity and level number variance [5], [6]. The spectral
rigidity, \(\Delta_3(L)\), is calculated for any real value \(L\) as:

@@

where \(\eta(\lambda)\) is the number of unfolded eigenvalues less than or equal to \(\lambda\),
\(\langle \cdot \rangle_c\) denotes the average with respect to all starting points \(c\), and where
\(A\) and \(B\) denote the slope and intercept, respectively, of the least squares fit of a straight
line to \(\eta(\lambda)\) on \([c, c+L]\) [5]. The level number variance, \(\Sigma^2(L)\), is
closely related to the spectral rigidity [6], and is calculated as:

@@

where \(\eta(L, c)\) is the number of unfolded eigenvalues in \([c, c+ L]\), and where \(c\), \(L\),
and \(\langle \cdot \rangle_c\) are as above [5]. Both the spectral rigidity and level number
variance are also computed and investigated in this study.

@@

The distribution of eigenvalues of large rectangular matrices with finite-variance, independent and
identically distributed (\textit{i.i.d.}) entries also converges to the Marchenko-Pastur (MP) distribution
[60], [61]:

@@

where \(\sigma^2\) is the true variance of the distribution of each i.i.d. entry (equal to 1 in the
case of correlation matrices), and where \(r = N / T\), for matrices of dimension \(N \times T\).
While real-world data will not, in general, be i.i.d., the extent to which spectra overlap with or
are well-described by the MP distribution has been used to guide analytical decisions regarding the
relevance of empirically observed eigenvalues. That is, since the majority of the eigenvalues of
random correlation matrices will fall within \([\lambda_{-}, \lambda_{+}]\), eigenvalues observed to
fall within this region are typically considered to be the result of noise [29]–[31], [62]. We
examine as a potential feature the proportion of eigenvalues that fall within this range, and call
this the noise ratio. In addition, we calculate a broader, shifted noise ratio, based on the
adjustments of [63] which takes into account the larger degree of correlation in empirical data.

\subsection{RMT Feature Extraction}
The entire 4-dimensional \(m \times n \times n \times p \times t\) scan array for each run,
excluding voxels with constant signals (which lead to undefined correlations), and excluding voxels
identified as non-brain via preprocessing, was flattened to a two-dimensional \(m \cdot n \cdot p
\times t\) matrix, and then the \(t - 1\) eigenvalues of that correlation matrix were computed. The
value of \(t\) for a scan can be inferred from the number of volumes in Table I. The various RMT
features mentioned above were then extracted via the empyricalRMT library [64] developed and made
publicly available as part of this study, as well as via specific code for this paper (which is
documented and available online [65]).

As both the spectral rigidity and level number variance can be highly sensitive to the choice of
smoothing procedure used in the unfolding method [58], [66]–[69], all unfolded features in this
study were calculated using polynomials of degree 5, 7, 9, 11, and 13 as the smoothing technique. As
much of the variance in the unfolded spectral statistics can be due to the largest eigenvalues
acting as anchor points during smoothing [66], [69], all unfolded features were also calculated
twice (in addition to each polynomial degree) to examine the effects of trimming on spectral
observables: once with only the largest eigenvalue omitted, and once with the largest 20 eigenvalues
omitted.

@@

A wide variety of standard, out-of-the-box classifiers available in the Python scikit-learn library
[70] were then employed to gauge the predictive value of the RMT features. To allow for a wide range
of flexibilities, classifiers employing k-nearest neighbours (KNN), with k values in \(\{1, 2, 3,
\dots, 10, 20\}\), logistic regression (LR), linear discriminant analysis (LDA), and a
support-vector machine (SVM) were employed. Default parameter values were used when possible (see
[65] for precise function calls used).

>>>>

For this paper’s analyses, brain extraction is necessary to limit analyses to the appropriate
voxels, motion correction is necessary to allow proper brain extraction, and both motion-correction
and slice-time correction simplify the interpretation of voxel-voxel correlations across voxel
locations. Whereas physiological noise regression and bias field correction would perhaps be
appropriate here, there is insufficient data to perform these steps for all datasets (Table II).
Thus, motion-correction was performed with FSL's MCFLIRT
(\cite{jenkinsonImprovedOptimizationRobust2002}), slicetime correction with the default method in
FSL, and brain extraction was performed with BET (\cite{smithFastRobustAutomated2002}). Due to
differences in the size of the REFLECT data, the higher-resolution resting-state data was
interpolated to have the same spatial dimensions as the task data.

@@

Two classes of features were examined in this study: those that require unfolding, and those that do
not. In order to compare the observed eigenvalue distribution to some theoretical predictions, the
eigenvalues must be unfolded (\cite{guhrRandommatrixTheoriesQuantum1998a,mehtaRandomMatrices2004}).
The details and motivation behind the unfolding procedure are well documented in
\cite{guhrRandommatrixTheoriesQuantum1998a}. Practically, the unfolding process can be viewed as a
smoothing and rescaling procedure where the originally observed spectrum \(\lambda_1 \le \dots \le
\lambda_n\) is smoothly mapped to the unfolded spectrum \(e_1, \dots, e_n\), such that \(\sum d_i/n
\approx 1\), if \(d_i = e_{i+1} - e_i \). For most empirical data, the true smoothing function is
not known, and so must be approximated
(\cite{guhrRandommatrixTheoriesQuantum1998a,mehtaRandomMatrices2004}). Typically, a polynomial is
chosen (\cite{abul-magdUnfoldingSpectrumChaotic2014}).

A number of statistics (or "spectral observables"; \cite{mehtaRandomMatrices2004}) can then be
computed from this unfolded spectrum. In the literature, the two most common points of comparison
for these unfolded features are typically the features of the Poisson or Gaussian Diagonal Ensemble
(GDE) and the GOE (see e.g.
\cite{santhanamStatisticsAtmosphericCorrelations2001,jalanUncoveringRandomnessSuccess2014,bandyopadhyayUniversalityComplexNetworks2007,agrawalQuantifyingRandomnessProtein2014,raiRandomnessPreservedPatterns2015,
sebaRandomMatrixAnalysis2003,wangSpectralPropertiesTemporal2015,wangRandomMatrixTheory2016}). The
GOE (or GDE) consists of real, symmetric matrices where entries (or only diagonal entries) are
normal random variables with zero mean and unit variance. When the empirically observed features are
close to those of the GOE, this is usually interpreted as indicating a high degree of randomness and
statistical independence in the data, and when the observed features are closer to those expected in
the GDE or Poisson case, this is interpreted as indicating a high degree of coherence or dependence
(\cite{santhanamStatisticsAtmosphericCorrelations2001,jalanUncoveringRandomnessSuccess2014,bandyopadhyayUniversalityComplexNetworks2007,agrawalQuantifyingRandomnessProtein2014,raiRandomnessPreservedPatterns2015,
sebaRandomMatrixAnalysis2003,wangSpectralPropertiesTemporal2015,wangRandomMatrixTheory2016,matharooSpontaneousBackpainAlters2020}).

The Nearest-Neighbours Spacings Distribution (NNSD) is the most basic RMT feature, and is the
distribution of the spacings (differences) of adjacent unfolded eigenvalues. Normally, the NNSD is
analysed visually via a histogram of the distribution of spacings, however it is also possible to
analyse it more formally by fitting the Brody distribution (\cite{brodyStatisticalMeasureRepulsion1973}):

@@

Whereas the NNSD and Brody parameter characterize the local properties of eigenvalues close on the
spectrum, the long-range properties relating eigenvalues distant on the spectrum can be
characterized by both the spectral rigidity and level number variance
(\cite{guhrRandommatrixTheoriesQuantum1998a,mehtaRandomMatrices2004}. The spectral rigidity,
\(\Delta_3(L)\), is calculated for any real value \(L\) as:

@@

where \(\eta(\lambda)\) is the number of unfolded eigenvalues less than or equal to \(\lambda\),
\(\langle \cdot \rangle_c\) denotes the average with respect to all starting points \(c\), and where
\(A\) and \(B\) denote the slope and intercept, respectively, of the least squares fit of a straight
line to \(\eta(\lambda)\) on \([c, c+L]\) \cite{guhrRandommatrixTheoriesQuantum1998a}. The level
number variance, \(\Sigma^2(L)\), is closely related to the spectral rigidity
(\cite{mehtaRandomMatrices2004}), and is calculated as:

@@

where \(\eta(L, c)\) is the number of unfolded eigenvalues in \([c, c+ L]\), and where \(c\), \(L\),
and \(\langle \cdot \rangle_c\) are as above (\cite{guhrRandommatrixTheoriesQuantum1998a}). Both the
spectral rigidity and level number variance are also computed and investigated in this study.

@@

The distribution of eigenvalues of large rectangular matrices with finite-variance, independent and
identically distributed (\textit{i.i.d.}) entries also converges to the Marchenko-Pastur (MP)
distribution (\cite{marcenkoDISTRIBUTIONEIGENVALUESSETS1967,yaoLargeSampleCovariance2015}):

@@

where \(\sigma^2\) is the true variance of the distribution of each i.i.d. entry (equal to 1 in the
case of correlation matrices), and where \(r = N / T\), for matrices of dimension \(N \times T\).
While real-world data will not, in general, be i.i.d., the extent to which spectra overlap with or
are well-described by the MP distribution has been used to guide analytical decisions regarding the
relevance of empirically observed eigenvalues. That is, since the majority of the eigenvalues of
random correlation matrices will fall within \([\lambda_{-}, \lambda_{+}]\), eigenvalues observed to
fall within this region are typically considered to be the result of noise
(\cite{veraartDenoisingDiffusionMRI2016,veraartDiffusionMRINoise2016,ulfarssonDimensionEstimationNoisy2008,macmahonCommunityDetectionCorrelation2015}).
We examine as a potential feature the proportion of eigenvalues that fall within this range, and
call this the noise ratio. In addition, we calculate a broader, shifted noise ratio, based on the
adjustments of \cite{almogUncoveringFunctionalSignature2019} which takes into account the larger
degree of correlation in empirical data.

\subsection{RMT Feature Extraction}
The entire 4-dimensional \(m \times n \times n \times p \times t\) scan array for each run,
excluding voxels with constant signals (which lead to undefined correlations), and excluding voxels
identified as non-brain via preprocessing, was flattened to a two-dimensional \(m \cdot n \cdot p
\times t\) matrix, and then the \(t - 1\) eigenvalues of that correlation matrix were computed. The
value of \(t\) for a scan can be inferred from the number of volumes in Table I. The various RMT
features mentioned above were then extracted via the empyricalRMT library [64] developed and made
publicly available as part of this study, as well as via specific code for this paper (which is
documented and available online (\cite{bergerderekm.DMBergerRandommatrixfmriV02020}).

As both the spectral rigidity and level number variance can be highly sensitive to the choice of
smoothing procedure used in the unfolding method
(\cite{abul-magdUnfoldingSpectrumChaotic2014,moralesImprovedUnfoldingDetrending2011,abueleninSpectralUnfoldingChaotic2018,fossionRandommatrixSpectraTime2013,abueleninEffectUnfoldingSpectral2012}),
all unfolded features in this study were calculated using polynomials of degree 5, 7, 9, 11, and 13
as the smoothing technique. As much of the variance in the unfolded spectral statistics can be due
to the largest eigenvalues acting as anchor points during smoothing
(\cite{moralesImprovedUnfoldingDetrending2011,abueleninEffectUnfoldingSpectral2012}), all unfolded
features were also calculated twice (in addition to each polynomial degree) to examine the effects
of trimming on spectral observables: once with only the largest eigenvalue omitted, and once with
the largest 20 eigenvalues omitted.

@@

A wide variety of standard, out-of-the-box classifiers available in the Python scikit-learn library
(\cite{pedregosaScikitlearnMachineLearning2011}) were then employed to gauge the predictive value of
the RMT features. To allow for a wide range of flexibilities, classifiers employing k-nearest
neighbours (KNN), with k values in \(\{1, 2, 3, \dots, 10, 20\}\), logistic regression (LR), linear
discriminant analysis (LDA), and a support-vector machine (SVM) were employed. Default parameter
values were used when possible (see \cite{bergerderekm.DMBergerRandommatrixfmriV02020} for precise
function calls used).
